---
title: "Presentation"
format:
  revealjs: 
    theme: sky
slide-number: true
editor: visual
---

## Introduction

Slide 1

## Materials and Methods

The raw data:

```{r}
# Read the TSV file
meta_untidy <- read.table("../data/01_meta_untidy.tsv",
                   sep = "\t",
                   header = TRUE)

library(reactable)
library(dplyr)
library(tidyverse)
library(tibble)
```

:::::: columns
::: {.column width="65%"}
```{r}
reactable(meta_untidy,
          sortable = TRUE,
          resizable = TRUE,
          height = "500px",
          theme = reactableTheme(
            style = list(fontSize = "20px")  # shrink text; default is 16px
          )
)

cat("dimension =", paste(dim(meta_untidy), collapse=" x "), "\n")
cat("Number of samples =", length(unique(meta_untidy$sample)), "\n")
```
:::

:::: {.column width="35%"}
::: {style="font-size: 0.7em;"}
But a tidy data set should follow these rules:

-   Every column contains exactly one variable.

-   Every row represents exactly one experimental unit (sample).

-   Every cell contains exactly one measurement.
:::
::::
::::::

## Materials and Methods


```{r}
meta_separated <- meta_untidy |>
  separate(characteristics, into = c("key", "value"), sep = ": ", extra = "merge")

meta_wide <- meta_separated |>
  pivot_wider(names_from = key, values_from = value)

meta_wide <- meta_wide |>
  mutate_all(~ ifelse(. == "NA", NA, .))

meta_wide <- meta_wide |> select(-tissue)

meta_clean <- meta_wide |>
  separate(group_day, into = c("group_temp", "day"), sep = "_", remove = TRUE) |>
  select(-group_temp)
```

::::: columns
::: {.column width="45%"}
::: {style="font-size: 0.7em;"}
Separate `characteristics` into key/value pairs
```r
meta_separated <- meta_untidy |>
  separate(characteristics, into = c("key", "value"), sep = ": ")
```
Convert long format to wide format
```r
meta_wide <- meta_separated |>
  pivot_wider(names_from = key, values_from = value)
```
Replace string `"NA"` with actual `NA`
```r
meta_wide <- meta_wide |>
  mutate_all(~ ifelse(. == "NA", NA, .))
```
Split `group_day` into two variables
```r
meta_clean <- meta_wide |>
  separate(group_day, into = c("group_temp", "day"), sep = "_", remove = TRUE) |>
  select(-group_temp)
```

:::
:::

::: {.column width="55%"}


```{r}
reactable(meta_clean,
          sortable = TRUE,
          resizable = TRUE,
          height = "600px",
          theme = reactableTheme(
            style = list(fontSize = "20px")  # shrink text; default is 16px
          )
)

cat("dimension =", paste(dim(meta_clean), collapse=" x "), "\n")
```
:::
:::::

## Materials and Methods

Slide 4

## Results

Slide 5

## Results

Slide 6

## Results

Slide 7

## Results

Slide 8

## Discussion

Slide 9

## Discussion

Slide 10
