---
title: "Exam - Augment Data: Group 03"
format:
  html:
    embed-resources: true
editor: visual
author: Mei Lin Verghese Law Kung Sam(s194685), Thelma Urena(s242552), Alejandra Caballero(s231912),Marco Bonafede(s243247), Carlos Sainz(s253695)
execute:
  warning: false
---

## Load libraries

```{r}
library(tidyverse)
library(tibble)
library(readr)
library(dplyr)
```

## Time to add our new genes

Before we had a metadata table, now we will add 15 new genes based on the paper (Pawlikowski JS, McBryan T, van Tuyn J, Drotar ME et al. Wnt signaling potentiates nevogenesis. Proc Natl Acad Sci U S A 2013 Oct 1;110(40):16009-14. PMID: [24043806](https://www.ncbi.nlm.nih.gov/pubmed/24043806 "Link to PubMed record"))

-   5 down regulated genes

    -   WIPF1, CD8A, CD2, ELK1 and CD3E

-   5 up regulated genes

    -   MAPK1, LFT, SOCS1, RELB and GNB4

-   5 non modified genes

    -   TRIM44, UGT2B7, TCP1, ORC1L and NEK1

### First, we load the gene expression matrix

```{r}
gene_expression <- read_tsv("../data/expression_data.tsv")
head(gene_expression)
```

We need to filter this table for the genes of interest previously mentioned only..

Oh no! we find out that we don't have the real gene name, but the illumina gene ID. Lucky for us, we have a table that will help us with that

### Then, we load the previous gene table

So we can find the real gene name based on the ILMN gene name

```{r}
setwd("..")
gene_names <- read_tsv("data/platform_table_clean.tsv")
show(gene_names)
```

### Gene expression data formatting

Join both tables, by renaming columns and merging

```{r}
#Rename so both have the same ILMN_ID
expression_data <- gene_expression |>
  rename(ILMN_ID = gene_ID)

annotation_data <- gene_names |>
  rename(ILMN_ID = ILMN_Gene)

#Merge the two tables based on ILMN_ID
merged_data <- expression_data |>
  left_join(annotation_data, by = "ILMN_ID")

head(merged_data)
```

However, we stated in the beginning that we are only interested in certain genes, so lets select those so it's easier to work with

```{r}
# List of genes of interest
gene_list <- c(
  "MAPK1","LFT","SOCS1","RELB","GNB4","B2M","IRAK4","WIPF1",
  "CD8A","CD2","ELK1","CD3E","TRIM44","UGT2B7","TCP1","ORC1L","NEK1"
)

# Check that the column Gene_ID exists in merged_data
if (!("Gene_ID" %in% colnames(merged_data))) {
  stop("The column 'Gene_ID' does not exist in merged_data.")
}

#Before we filter the data, we want to save the unfiltered
merged_data_unfiltered = merged_data

# Filter the merged data by the gene list
merged_data_clean <- merged_data %>%
  dplyr::filter(Gene_ID %in% gene_list)

# Display the result
print(merged_data_clean)
```

Some of this gene names have more than 1 illumina gene identifier, so lets obtain an average gene expression for these gene names

```{r}
#Keep only genes of interest
merged_data <- merged_data_clean |>
  filter(!is.na(Gene_ID))

#Average values with same Gene_ID
clean_data <- merged_data |>
  group_by(Gene_ID) |>
  summarise(across(starts_with("GSM"), mean, na.rm = TRUE)) |>
            ungroup()

#Now we have only the data we are interested in 
length(unique(clean_data$Gene_ID))
show(clean_data)
```

Ok, now the expression_data is in format!

Now, lets merge the original metadata and gene expression based on sample_id

```{r}
#First, pivot gene expression
second_clean <- clean_data |>
  pivot_longer(
    cols = starts_with("GSM"),
    names_to = "sample",
    values_to = "expression"
  )

show(second_clean)
```

Now we can merge them!

```{r}

meta_clean = read_tsv(file.path("../data/02_meta_clean.tsv"))

merged_final <- second_clean |>
  left_join(meta_clean, by = "sample")
```

### Final data visualization

```{r}
show(merged_final)
summary(merged_final)

write_tsv(x = merged_final,
          file = file.path("../data/03_merged_final.tsv"))
```

### Augmenting the unfiltered data

We make the same changed to the unfiltered data, as we did to the filtered data

```{r}
clean_data_unfiltered <- merged_data_unfiltered |>
  group_by(Gene_ID) |>
  summarise(across(starts_with("GSM"), mean, na.rm = TRUE)) |>
            ungroup()

second_clean_unfiltered <- clean_data_unfiltered |>
  pivot_longer(
    cols = starts_with("GSM"),
    names_to = "sample",
    values_to = "expression"
  )

merged_final_unfiltered <- second_clean_unfiltered |>
  left_join(meta_clean, by = "sample")
```

Then we save it

```{r}
write_tsv(x = merged_final_unfiltered,
          file = file.path("../data/03_merged_final_unfiltered.tsv"))
```

### Augment Phenotype data

Read phenotype data

```{r}
pheno_data <- read_tsv("../data/phenotype_data.tsv")
```

Detect which rows has the word 'Control' in the title column. If they do, for the new column 'control', equal TRUE, if not, equal FALSE.

```{r}
pheno_data_control <- pheno_data |> 
  mutate(control = str_detect(title, 
                              '(Control)'))  |> 
  mutate(control = factor(control, 
                          levels = c(TRUE, FALSE)))  |> 
  group_by(control)
```

Remove all column except geo_accession and control

```{r}
pheno_data_control <- pheno_data_control  |> 
  select(geo_accession, control)
```

### Augmenting for shiny

We want to use merged_final_unfiltered for the shiny. For this, we remove unnecessary columns

```{r}
merged_unfiltered <- merged_final_unfiltered |> 
  select(Gene_ID, 
         sample, 
         expression)
```

Merge merged_unfiltered and pheno_data_control

```{r}
merged_pheno_control <- merged_unfiltered |> 
  left_join(y = pheno_data_control,
            by = join_by(sample == geo_accession))
```

We want to get the mean gene expression for every unique Gene ID and control. For this we need to group the data by Gene ID and control.

```{r}
#Get list of distinct gene_ID
Gene_ID_distinct <- merged_pheno_control |> 
  distinct(Gene_ID)

#Add control and gene_id as factors and groups
merged_pheno_control <- merged_pheno_control |> 
  mutate(control = factor(control, 
                          levels = c(TRUE, FALSE)),
         Gene_ID = factor(Gene_ID,
                          Gene_ID_distinct$Gene_ID))  |> 
  group_by(control,
           Gene_ID)
```

We then make a new tibble with only the Gene ID, control and mean gene expression

```{r}
#Get the mean gene expression for each gene
merged_mean <- merged_pheno_control |> 
  summarize(mean = mean(expression))
```

We then want to get the difference between mean gene expression for if control is true or false

```{r}
#Pivot wide, turning control into two columns with values from mean
merged_mean <- merged_mean |> 
  pivot_wider(names_from = control,
              values_from = mean)

#Add column with difference in gene expression between control and non-control
#For each gene
merged_mean <- merged_mean |> 
  mutate(diff = `FALSE`-`TRUE`)
```

Then we join merged_mean and merged_pheno_control

```{r}
merged_pheno_control <- merged_pheno_control |> 
  left_join(y = merged_mean,
            by = join_by(Gene_ID))

merged_pheno_control <- merged_pheno_control |> 
  ungroup()
```

Then we pivot merged_pheno_control long, so we have one column with if control is true or false and another with mean gene expression

```{r}
merged_pheno_control <- merged_pheno_control |> 
  select(Gene_ID, 
         `TRUE`, 
         `FALSE`, 
         diff)

merged_pheno_control <- merged_pheno_control |> 
  pivot_longer(col = c(`TRUE`, `FALSE`), 
               names_to = "control", 
               values_to = "mean_gene_expression")
```

We then remove repeated rows and save it

```{r}
merged_pheno_control <- merged_pheno_control |> 
  distinct()


write_tsv(x = merged_pheno_control, 
          file = "../data/03_merged_shiny.tsv")

```
