---
title: "Exam - Analysis Data Part 1: Group 03"
format:
  html:
    embed-resources: true
editor: visual
author: Mei Lin Verghese Law Kung Sam(s194685), Thelma Urena(s242552), Alejandra Caballero(s231912),Marco Bonafede(s243247), Carlos Sainz(s253695)
execute:
  warning: false
---

# Load Libraries

```{r}
library(tidyr)
library(tidyverse)
library(readr)
library(dplyr)
library(ggplot2)
library(broom)
library(purrr)
library(stringr)
library(forcats)
library(ggrepel)
```

# Load Data

```{r}
# Load the merged, tidy data
merged_final <- read_tsv("../data/03_merged_final.tsv")

# Take a quick look at the structure
glimpse(merged_final)
show(merged_final)
```

# Ensure variables are correct type and filter for Day 1 analysis, also check group count

```{r}
merged_prepared <- merged_final |>
  mutate(
    disease_status = factor(`disease status`),
    day = factor(day),
    gender = factor(gender),
    neutrophil_proportion = as.numeric(`neutrophil proportion`)
  )

# Filter to day 1 samples only
merged_day1 <- merged_prepared |>
  filter(day == "D1")

# Check how many samples per disease group at day 1
merged_day1 |>
  count(disease_status)
```

<<<<<<< HEAD
# Boxplots of gene expression by disease status on day 1 for every gene
```{r fig.width=10, fig.height=7}
#So we can visualize plots
options(bitmapType = "cairo")

merged_day1 |>
  ggplot(aes(
    x = disease_status,
    y = expression,
    fill = disease_status
  )) +
  geom_boxplot(outlier.size = 0.8, alpha = 0.8) +
  facet_wrap(~ Gene_ID, scales = "free_y", nrow = 3) + 
  scale_fill_manual(
    values = c(
      "healthy" = "#4DAF4A",
      "sepsis nonsurvivor" = "#E41A1C",
      "sepsis survivor" = "#377EB8"
    )
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = "bottom"
  ) +
  labs(
    title = "Gene expression (Day 1) by disease status ",
    x = "Disease status",
    y = "Expression",
    fill = "Disease status"
  )
```

# Now lets exclude healthy and take a look on survivors vs non survivors
```{r}
# Filter to only sepsis samples
sepsis_day1 <- merged_prepared |>
  filter(day == "D1" & disease_status != "healthy")

# Check group counts
sepsis_day1 |> count(disease_status)
```

<<<<<<< HEAD
# Same plot (gene expression by disease status) but for survivors vs non survivors
```{r fig.width=10, fig.height=7}
survival_boxplots <- ggplot(sepsis_day1, aes(
    x = disease_status,
    y = expression,
    fill = disease_status
)) +
  geom_boxplot(outlier.size = 0.8, alpha = 0.9) +
  facet_wrap(~ Gene_ID, scales = "free_y", nrow = 3) +
  scale_fill_manual(
    values = c(
      "sepsis nonsurvivor" = "#E41A1C",
      "sepsis survivor" = "#377EB8"
    )
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = "bottom",
    strip.text = element_text(size = 12)
  ) +
  labs(
    title = "Gene expression on day 1 by survival status",
    x = "Survival status",
    y = "Expression",
    fill = "Survival status"
  )

ggsave(
  filename = "~/projects/R_project/doc/survival_boxplots_day1.png",
  plot = survival_boxplots,
  width = 10,
  height = 7,
  dpi = 300
)
```
# Calculating mean and standard deviation for survivors and non survivors for each gene

```{r}
survival_summary <- sepsis_day1 |>
  group_by(Gene_ID, disease_status) |>
  summarise(
    mean = mean(expression),
    sd = sd(expression),
    n = n(),
    .groups = "drop"
  )

survival_summary
```

# T test on mean expression in survivors vs non survivors

```{r}

survival_tests <- sepsis_day1 |>
  group_by(Gene_ID) |>
  group_modify(~{
    test <- t.test(expression ~ disease_status, data = .x)
    tidy(test)
  }) |>
  ungroup()

#renamed for more clarity
survival_tests_renamed <- survival_tests |>
  rename(
    mean_difference   = estimate,
    mean_nonsurvivor  = estimate1,
    mean_survivor     = estimate2
  )


survival_tests_renamed
```

# Combining summary statistics with t test results for each gene 

```{r}
#pivot summary to wide form

summary_wide <- survival_summary |>
  select(Gene_ID, disease_status, mean) |>
  mutate(
    disease_status = str_replace_all(disease_status, " ", "_")
  ) |>
  pivot_wider(
    names_from = disease_status,
    values_from = mean,
    names_prefix = "mean_"
  )

#join with t-test results
survival_results <- survival_tests_renamed |>
  left_join(summary_wide, by = "Gene_ID") |>
  mutate(
    diff_nonsurvivor_minus_survivor =
      mean_sepsis_nonsurvivor - mean_sepsis_survivor,
    significant_p = p.value < 0.05
  ) |>
  arrange(p.value)

survival_results
```

# Final plot for expression difference and significance (based on adjusted p value)

```{r}
expression_diff_plot <- ggplot(survival_results, aes(
    x = reorder(Gene_ID, diff_nonsurvivor_minus_survivor),
    y = diff_nonsurvivor_minus_survivor,
    fill = significant_p
)) +
  geom_col() +
  coord_flip() +
  theme_bw() +
  labs(
    title = "Expression difference between the two groups",
    x = "Gene",
    y = "Mean difference",
    fill = "Significant (p < 0.05)"
  )

ggsave(
  filename = "~/projects/R_project/doc/expression_difference_day1.png",
  plot = expression_diff_plot,
  width = 10,
  height = 7,
  dpi = 300
)

```
INTERPRETATION:

This plot shows the difference in mean gene expression between sepsis nonsurvivors and survivors on Day 1, where positive values indicate higher expression in nonsurvivors and negative values indicate higher expression in survivors. Several genes, such as SOCS1, ELK1, GNB4, CD8A, and MAPK1, display noticeable expression differences between the two groups. Bars colored in blue represent genes whose raw p value is below 0.05, indicating a statistically significant difference in expression between survivors and nonsurvivors at this threshold. This visualization highlights which genes show the strongest separation between clinical outcomes early in sepsis and could provide useful features for predictive models aimed at identifying high-risk patients who may benefit from earlier or more intensive care.

THELMA'S part

```{r}
show(merged_final)
```

Understanding better the data

Relation between age, sex and disease status

```{r}

ggplot(merged_final, aes(x = gender, y = `age (years)`, fill = `disease status`)) +
  geom_boxplot(alpha = 0.7) +
  theme_bw() +
  labs(
    title = "Age distribution by sex and disease status",
    x = "Sex", y = "Age", fill = "Disease Status"
  ) +
  theme(plot.title = element_text(hjust = 0.5))

```

Most common site of infection

```{r}
ggplot(merged_final, aes(x = fct_infreq(`site of infection`), fill = `disease status`)) +
  geom_bar(position = "dodge") +
  coord_flip() +
  theme_bw() +
  labs(
    title = "Cases by infection site and disease status",
    x = "Site of Infection",
    y = "Number of cases",
    fill = "Disease Status"
  ) +
  theme(plot.title = element_text(hjust = 0.5))


```

```{r}
ggplot(merged_final, aes(x = Gene_ID, y = expression, fill = Gene_ID)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  geom_boxplot(width = 0.1, outlier.size = 0.8, alpha = 0.8) +
  theme_bw() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5)
  ) +
  labs(
    title = "Gene expression distribution per gene",
    y = "Expression level",
    x = "Gene ID"
  )
```

Expression per gene across disease status

```{r}

ggplot(merged_final, aes(x = `disease status`, y = expression, fill = `disease status`)) +
  geom_violin(alpha = 0.6) +
  geom_boxplot(width = 0.15, outlier.size = 0.5, alpha = 0.8) +
  facet_wrap(~ Gene_ID, scales = "free_y") +
  theme_bw() +
  labs(title = "Expression per gene across disease status",
       x = "Disease status", y = "Expression") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1))

```

Calculating difference of mean in gene expression - sepsis nonsurvivor vs healthy positive = overexpressed () negative = underexpressed

```{r}
# calculating mean by gene and disease condition
expr_diff <- merged_final %>%
  group_by(Gene_ID, `disease status`) %>%
  summarise(mean_expr = mean(expression, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = `disease status`, values_from = mean_expr) %>%
# Dealing with NA
  mutate(
    diff_NS_vs_H = `sepsis nonsurvivor` - healthy,
    direction = case_when(
      is.na(diff_NS_vs_H) ~ "no_data",
      diff_NS_vs_H > 0 ~ "overexpressed",
      diff_NS_vs_H < 0 ~ "underexpressed",
      TRUE ~ "no_change"
    )
  ) %>%
  arrange(desc(diff_NS_vs_H))

# plot
ggplot(expr_diff, aes(x = fct_reorder(Gene_ID, diff_NS_vs_H), y = diff_NS_vs_H, fill = direction)) +
  geom_col() +
  scale_fill_manual(values = c(
    "overexpressed" = "#D55E00",  
    "underexpressed" = "#0072B2", 
    "no_change" = "grey80",
    "no_data" = "grey95"
  )) +
  coord_flip() +
  theme_bw() +
  labs(
    title = "Difference in expression (Nonsurvivor - Healthy)",
    x = "Gene",
    y = "Expression difference (NS - Healthy)",
    fill = "Direction"
  ) +
  theme(plot.title = element_text(hjust = 0.5))

```

Volcano plot

```{r}
groupA <- "healthy"
groupB <- "sepsis nonsurvivor"
pseudocount <- 1e-6
alpha <- 0.05

# Data prep
volcano_simple <- merged_final %>%
  filter(`disease status` %in% c(groupA, groupB)) %>%
  group_by(Gene_ID) %>%
  summarise(
    mean_A = mean(expression[`disease status` == groupA], na.rm = TRUE),
    mean_B = mean(expression[`disease status` == groupB], na.rm = TRUE),

    # fold-change
    log2FC = log2((mean_B + pseudocount) / (mean_A + pseudocount)),

    # p-value
    p.value = {
      x <- expression[`disease status` == groupA]
      y <- expression[`disease status` == groupB]
      if (sum(!is.na(x)) >= 2 && sum(!is.na(y)) >= 2) {
        tryCatch(t.test(y, x)$p.value, error = function(e) NA_real_)
      } else {
        NA_real_
      } 
    },
    .groups = "drop"
  ) %>%
  mutate(
    padj = p.adjust(p.value, method = "fdr"),
    sig = case_when(
      is.na(padj) ~ "no_data",
      padj < alpha & log2FC > 0 ~ "overexpressed",
      padj < alpha & log2FC < 0 ~ "underexpressed",
      TRUE ~ "not_significant"
    )
  )

# Volcano Plot
ggplot(volcano_simple, aes(x = log2FC, y = -log10(padj))) +
  geom_point(aes(color = sig), alpha = 0.7, size = 2) +
  scale_color_manual(values = c(
    "overexpressed" = "#D55E00",
    "underexpressed" = "#0072B2",
    "not_significant" = "grey70",
    "no_data" = "grey95"
  )) +
  geom_text_repel(
    data = volcano_simple %>% filter(sig %in% c("overexpressed", "underexpressed")),
    aes(label = Gene_ID),
    size = 3,
    max.overlaps = 20
  ) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "darkgrey") +
  geom_hline(yintercept = -log10(alpha), linetype = "dashed", color = "darkgrey") +
  theme_bw() +
  labs(
    title = paste(groupB, "vs", groupA),
    x = "log2 Fold Change (B / A)",
    y = "-log10(FDR p-value)",
    color = "Significance"
  ) +
  theme(plot.title = element_text(hjust = 0.5))


```

Analyzing the gene expression during the 5 days and its relation with the disease status

```{r}
ggplot(merged_final, aes(x = day, y = expression, color = `disease status`)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", se = FALSE) +
  facet_wrap(~ Gene_ID, scales = "free_y") +
  theme_bw()

```

Gene expression vs neutrophil proportion

Only with genes showing more variance in expression - is there any co-relation?

```{r}

# Choosing genes with more variance in expression 
top_genes <- merged_final %>%
  group_by(Gene_ID) %>%
  summarise(var_expr = var(expression, na.rm = TRUE)) %>%
  arrange(desc(var_expr)) %>%
  slice_head(n = 12) %>%
  pull(Gene_ID)

df_facet <- merged_final %>% filter(Gene_ID %in% top_genes)

ggplot(df_facet, aes(x = `neutrophil proportion`, y = expression)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, color = "red", linetype = "dashed") +
  facet_wrap(~ Gene_ID, scales = "free_y") +
  theme_bw() +
  labs(title = "Expression vs neutrophils — selected genes",
       x = "Neutrophil proportion", y = "Expression") +
  theme(plot.title = element_text(hjust = 0.5))

```
